Shaping strategies and standardisation of faceted spheroids from Qianshangying (North China)

### 2025-07-02 analysis/paper/paper.qmd
updates_material_discussion

The pursuit of perfect geometric forms—particularly the ideal sphere—appears to transcend time, persisting from antiquity to the present. Spheroids may be kind of stone artifacts that embody this aspiration toward geometric perfection, reflecting Paleolithic humans’ spatial and navigational cognition, and their grasp of abstract concepts such as ''3D symmetry'' ([@Muller2023; @clark1955; @wynn2002]). This cognitive awareness may have emerged even earlier than the production of handaxes ([@beyene2013; @shipton2019; @white2018]). Consequently, the study of spheroids—including their manufacturing techniques and morphological characteristics—has become a key issue in interpreting early human technological behaviors and cognitive capacities ([@cabanès2022]).

@@

Although these artifact types are frequently grouped under the category “PSB” (or “PSSB” when subspheroids are included) and are commonly interpreted as volumetrically reduced artifacts organized around a central point, this morphology-based classification has led to several interpretive challenges. Artifacts identified as spheroids in one study may be classified as bolas or attributed to entirely different technological categories in another. Furthermore, research has shown that these forms may not represent a unified *chaîne opératoire*. Differences in raw material selection suggest divergent technological strategies and potentially distinct functional purposes ([@deweyer2017; @jones1994]).

At present, the technological category and function of spheroids remain central topics of debate. Broadly, spheroids have been interpreted as hammerstones ([@mussi2025; @schick1994], cores ([@sahnouni1997]), or projectiles ([@leakey1979]). Closely tied to this debate is the question of their production strategy, which remains unclear. Experimental replication and site-specific diacritical analyses have produced divergent interpretations. One view suggests that spheroids emerged unintentionally as the by-products of core exhaustion, which represents a low-effort strategy aimed at flake production ([@sahnouni1997]). Another perspective holds that spheroids were naturally produced through high-intensity battering ([@schick1994]; \[\@mora2005\]; [@clark1955]). In contrast, a third view proposes that spheroids reflect a conceptual template and were produced through a deliberate shaping strategy ([@tittonSubspheroidsLithicAssemblage2020; @Muller2023; @texier2014]).

Spheroids are a widely distributed type of stone artifact, appearing across a broad temporal and geographic range—from the Oldowan and Acheulean to the Middle Paleolithic—and in diverse archaeological assemblages ([@willoughby1985; @Cabanes2024]). This wide distribution is matched by considerable technological and morphological variability, which has fueled ongoing debates concerning their methods of production and functional significance. Site-specific studies are therefore essential for reconstructing the spatial and temporal distribution of spheroids, as well as for identifying their distinct technological traits and functions. Such analyses provide a foundation for situating spheroids within broader discussions of early human technological behavior.

In this study, we analyzed material from the Middle Pleistocene Qianshangying site (QSY) in northern China. Among an assemblage previously classified as a simple core-flake industry, we identified a group of highly standardized spheroids and subspheroids. we observed notable differences between these artifacts and multifacial cores or polyhedrons, which are stone artifacts through typical flaking strategies. We proposed that their concept and manufacture reflect a distinct technological tradition, separate from conventional flake-based technology.

We conducted quantitative morphological and technological analyses, including the novel spherical harmonic analysis, on the spheroids recovered from the QSY site. Our goal was to evaluate whether these spheroids exhibited standardized morphology and production patterns, and whether their tendency toward spherical form resulted from deliberate and systematic reduction processes. These analyses aimed to clarify the technological strategy behind spheroid production at QSY, within the broader context of Middle Pleistocene lithic practices in northern China.

@@

The QSY lies in the Nihewan Basin, a region in northern China situated between the Chinese Loess Plateau and the Inner Mongolian Plateau. The basin preserves extensive Quaternary fluvio-lacustrine and loess deposits ([@li2000; @yuan2009]), hosts the highest concentrations of Pleistocene Paleolithic sites in East Asia, and is especially significant for preserving evidence of early hominin activities outside Africa dating back over one million years ([@yang2020; @yang2017; @pei2019; @pei2017; @zhu2001; @zhu2004]). In recent years, nearly 20 Middle Pleistocene sites—including Jijiazhuang and Caijiagou—have been identified in the southeastern part of the basin ([@yuwei2023; @ye2024]), within the Yuxian Sub-basin, thereby completing the regional cultural sequence of the Nihewan Basin.

Qianshangying site complex consisting of 4 localities from QSY-A to QSY-D, discovered in 2015, is among the region’s most artifact-rich archaeological sites. Geomorphological and sedimentological evidence indicates that the stratigraphy at the site records a full cycle of lake expansion, retreat, re-expansion, and eventually disappearance of the ancient Nihewan lake, and hominin occupation occurred in a marginal lacustrine setting during a low lake-level episode.

The stratigraphic context of QSY indicates a short-term occupation resulting from a single hominin activity event. The assemblage appears to be largely in situ, exhibiting minimal disturbance after deposition and well-preserved patterns of technological combination. ESR dating of the cultural layer places hominin activity at approximately 429 ± 39 ka.

A total of 957 stone artifacts were recovered from four localities at the QSY site, with QSY-A and QSY-B yielding the most abundant assemblages. Following a widely used techno-typological framework (Isaac, 1981, 1986; de la Torre and Mora, 2005, 2018; de la Torre et al., 2011; Pei et al., 2017), the assemblage was classified into four general categories: flaked pieces (n = 247, 25.9%), detached pieces (n = 686, 71.9%), pounded pieces (n = 4, 0.4%), and unmodified materials (n = 17, 1.8%).

@@

Based on this diacritical analysis and scar orientation analysis, we can summarize the technical strategy employed in the production of faceted spheroids in QSY. This strategy consists of a series of sequential phases (@fig-shaping-process): a potential rough-out phase, volume exploitation aimed at forming a cubic shape, surface exploitation to approximate a spheroidal geometry, edge refinement, and potentially the repetition of earlier phases. The existence of a repetitive phase is supported by the observation that while the original cobble blanks in QSY display considerable size variation, the finished spheroids—although some are relatively large—mostly cluster around a diameter of 8–10 cm. This suggests that the production of spheroids may have involved multiple cycles of reduction, and further implies that a diameter of 8–10 cm was considered an optimal or intended size in the knapping process. <!-- BM: maybe move to discussion -->

<!-- consider to add a visualisation, such as a histogram of the diameter, to support this claim, maybe above we can have a panel of plots of several basic attributes just to introduce the assemblages to the reader, then refer to it here. Maybe also some representative photos-->

@@

>>>>

Spheroids are a widely distributed type of stone artifact, appearing across a broad temporal and geographic range—from the Oldowan and Acheulean to the Middle Paleolithic—and in diverse archaeological assemblages.([@willoughby1985; @Cabanes2024]).

@@

Although these artifact types are frequently grouped under the category “PSB” of Polyhedron, Spheroid and Bolas (or “PSSB” when subspheroids are included) and are commonly interpreted as volumetrically reduced artifacts organized around a central point, this morphology-based classification has led to several interpretive challenges. Artifacts identified as spheroids in one study may be classified as bolas or attributed to entirely different technological categories in another. Furthermore, research has shown that these forms may not represent a unified *chaîne opératoire* : differences in raw material selection suggest divergent technological strategies and potentially distinct functional purposes ([@deweyer2017; @jones1994]).

At present, the technological category and function of spheroids remain central topics of debate. Broadly, spheroids have been interpreted as hammerstones ([@mussi2025; @schick1994], cores ([@sahnouni1997]), or projectiles ([@leakey1979]). Closely tied to this debate is the question of their production strategy, which remains unclear. Experimental replication and site-specific diacritical analyses have produced divergent interpretations. One view suggests that spheroids emerged unintentionally as the by-products of core exhaustion, which represents a low-effort strategy aimed at flake production ([@sahnouni1997]). Another perspective holds that spheroids were naturally produced through high-intensity battering ([@schick1994]; \[\@mora2005\]; [@clark1955]). In contrast, a third view proposes that spheroids reflect a conceptual template and were produced through a deliberate shaping strategy ([@tittonSubspheroidsLithicAssemblage2020; @Muller2023; @texier2014]). \[BM: yes, this is great, good job, this is a very important paragraph for the introduction of the paper. We might need to expand it a bit\]

Rather than reflecting a single functional or technological model, these varying interpretations may instead capture regional variability shaped by differences in raw materials, morphologies, and reduction strategies. As [@Cabanes2024] said, spheroids may not represent a homogenous category. \[BM: you have already mentioned debates in the previous paragraph, so it's seems a bit odd to mention debates again here without continuity with the previous paragraph. Maybe these two sentences are better as the opening sentences for the paper, setting the scene for more information about the debates\] This highlights the need to move beyond generalized typological categories and instead engage more directly with local contexts and specific assemblages. Site-specific studies are therefore essential for reconstructing the spatial and temporal distribution of spheroids, as well as for identifying their distinct technological traits and functions.

In this study, we analyzed material from the Middle Pleistocene Qianshangying site (QSY) in northern China. Among an assemblage previously classified as a simple core-flake industry, we identified a group of highly standardized spheroids and subspheroids. we observed notable differences between these artifacts and multifacial cores or polyhedrons, which are stone artifacts through typical flaking strategies. Our goal was to evaluate whether these spheroids exhibited standardized morphology and production patterns, and whether their tendency toward spherical form resulted from deliberate and systematic reduction processes. based on the results, further to answer the question of whether the concept and manufacture of these spheroids represent a distinct technological tradition, separate from conventional flake-based strategies.

We conducted quantitative morphological and technological analyses of the spheroids recovered from the Qianshangying (QSY) site, including the novel application of spherical harmonic analysis to a Paleolithic assemblage from East Asia. These analyses help clarify the technological strategy behind spheroid production at QSY, within the broader context of Middle Pleistocene lithic practices in northern China.

@@

QSY lies in the Nihewan Basin, a region in northern China situated between the Chinese Loess Plateau and the Inner Mongolian Plateau. The basin preserves extensive Quaternary fluvio-lacustrine and loess deposits ([@li2000; @yuan2009]), hosts the highest concentrations of Pleistocene Paleolithic sites in East Asia, and is especially significant for preserving evidence of early hominin activities outside Africa dating back over one million years ([@yang2020; @yang2017; @pei2019; @pei2017; @zhu2001; @zhu2004]). In recent years, nearly 20 Middle Pleistocene sites—including Jijiazhuang and Caijiagou—have been identified in the southeastern part of the basin ([@yuwei2023; @ye2024]), within the Yuxian Sub-basin, thereby completing the regional cultural sequence of the Nihewan Basin. \[BM: great detail about the context, can we make a figure that includes a map and a photo of the site?\]

The QSY site complex consists of four localities from QSY-A to QSY-D, discovered in 2015. It is among the region’s most artifact-rich archaeological site. Geomorphological and sedimentological evidence indicates that the stratigraphy at the site records a full cycle of lake expansion, retreat, re-expansion, and eventually disappearance of the ancient Nihewan lake, and hominin occupation occurred in a marginal lacustrine setting during a low lake-level episode \[BM: citation needed\]. The stratigraphic context of QSY indicates a short-term occupation resulting from a single hominin activity event. The assemblage appears to be largely in situ, exhibiting minimal disturbance after deposition and well-preserved patterns of technological combination. ESR dating of the cultural layer places hominin activity at approximately 429 ± 39 ka \[BM: citation needed\].

A total of 955 stone artifacts were recovered from four localities at the QSY site. Following a widely used techno-typological framework (Isaac, 1981, 1986; de la Torre and Mora, 2005, 2018; de la Torre et al., 2011; Pei et al., 2017), the assemblage was classified into four general categories: flaked pieces (n = 247, 26.0%), detached pieces (n = 686, 71.8%), pounded pieces (n = 4, 0.4%), and unmodified materials (n = 17, 1.8%). \[BM: is it possible to use inline R code for these numbers?\] Among these, 12 spheroids and subspheroids were identified, accounting for 1.3% of the total lithic assemblage.

Raw materials at QSY were classified into four categories following [@pei2002] and [@pei2017]: lava, siliceous dolomite, chert, and others. Lava rock dominates the assemblage (n=637, 66.7%), followed by siliceous dolomite (n=209, 21.9%), with chert and other materials comprising the remainder (n=53, 5.5% and n=56, 5.9% respectively). It seems that spheroids show no strong raw material preference. Of the 12 identified spheroids, 75.0% are lava (n = 9), 16.7% are siliceous dolomite (n = 2), and 8.3% is quartzite (n=1, classified under "other"), roughly mirroring the assemblage’s overall composition except for the absence of chert. However, when compared to cores, which are 66.0% lava (n = 62), 22.3% siliceous dolomite (n = 21), 4.3% chert (n = 4), and 7.4% other (n = 7), spheroids show a higher frequency of lava use, suggesting a possible preference or selection difference in their production. (pie chart in need?)

```{r, echo=FALSE}
#| label: fig-core-spheroid-size
#| fig-cap: "box plot showing size of cores and spheroids"
#| 
# 读取表格
library(tidyverse)
library(readxl)
library(here)
library(dplyr)

file_path <- here("analysis/data/derived_data/cores_spheroids_size.xlsx")
data <- read_excel(file_path)

summary_vars <- data %>%
  select(core_length, core_width, core_thickness, core_weight,
         spheroid_length, spheroid_width, spheroid_thickness, spheroid_weight,
          multi_poly_length, multi_poly_width, multi_poly_thickness)

summary_stats <- summary_vars %>%
  summarise(across(everything(),
                   list(mean = ~mean(.x, na.rm = TRUE),
                        sd = ~sd(.x, na.rm = TRUE)),
                   .names = "{.col}_{.fn}"))

long_data <- data %>%
  select(core_length, core_width, core_thickness,
         spheroid_length, spheroid_width, spheroid_thickness,
         multi_poly_length, multi_poly_width, multi_poly_thickness) %>%
  pivot_longer(
    everything(),
    names_to = "type_metric",
    values_to = "value"
  ) %>%
  extract(type_metric, into = c("artifact", "metric"),
          regex = "^(core|spheroid|multi_poly)_(length|width|thickness)") %>%
  filter(!is.na(value)) %>%
  mutate(
    artifact = recode(artifact,
                      "core" = "Core",
                      "spheroid" = "Spheroid",
                      "multi_poly" = "Multi-poly Core"),
    metric = factor(metric, levels = c("length", "width", "thickness"))
  )

# 画图
ggplot(long_data, aes(x = metric, y = value, fill = artifact)) +
  geom_boxplot(position = position_dodge(0.8)) +
  labs(
    title = "Comparison of Core, Spheroid, and Multi-poly Core Dimensions",
    x = "Dimension",
    y = "Measurement (mm)",
    fill = "Artifact"
  ) +
  scale_fill_manual(values = c("Core" = "#66c2a5", 
                               "Spheroid" = "orange", 
                               "Multi-poly Core" = "lightgray")) +
  theme_minimal(base_size = 14)

In terms of dimensions, the average length, width, thickness, and weight of spheroids were `r round(summary_stats$spheroid_length_mean, 1)` mm (SD = `r round(summary_stats$spheroid_length_sd, 1)`), `r round(summary_stats$spheroid_width_mean, 1)` mm (SD = `r round(summary_stats$spheroid_width_sd, 1)`), `r round(summary_stats$spheroid_thickness_mean, 1)` mm (SD = `r round(summary_stats$spheroid_thickness_sd, 1)`), and `r round(summary_stats$spheroid_weight_mean, 1)` g (SD = `r round(summary_stats$spheroid_weight_sd, 1)`), respectively, as shown in @fig-core-spheroid-size. Two spheroids showed outlier values in size, being either unusually large or small, while the rest exhibited relatively concentrated size distributions. In contrast, cores displayed greater variability in size, with standard deviations of `r round(summary_stats$core_length_sd, 1)` mm, `r round(summary_stats$core_width_sd, 1)` mm, and `r round(summary_stats$core_thickness_sd, 1)` mm for length, width, and thickness, respectively. The polyhedrons and multifacial cores showed less variation than the overall core group, but their dimensional variability remained substantially greater than that of the spheroids.Additionally, spheroids tend to have similar values across length, width, and thickness, reflecting their overall symmetry. In contrast, cores, including polyhedrons and multifacial cores show marked differences among these dimensions, indicating more irregular and elongated forms.

@@

@@

## Do Qianshangying Spheroids represent of a Shaping Strategy?

According to the results of morphological and technological analysis, we can summarize the technical strategy employed in the production of faceted spheroids in QSY. This strategy consists of a series of sequential phases (@fig-shaping-process): a potential rough-out phase, volume exploitation aimed at forming a cubic shape, surface exploitation to approximate a spheroidal geometry, edge refinement, and potentially the repetition of earlier phases. The existence of a repetitive phase is supported by the observation that while the original cobble blanks in QSY display considerable size variation, the finished spheroids, although two of them are relatively large or small, mostly cluster around a diameter of 6–9 cm (@fig-core-spheroid-size). This suggests that the production of spheroids may have involved multiple cycles of reduction, and further implies that a diameter of 6–9 cm was considered an optimal or intended size in the knapping process.

Therefore, it can be inferred that the QSY hominin deliberately employed a standardized production approach to create spheroids with consistent and formalized shapes. The technological strategy reflected in the QSY facted spheroids includes the use of conceptual template, standardized production, and the achievement of a target product, which together demonstrate that this was a shaping strategy. This stands in contrast to flaking strategies, where the primary objective is the removal of usable flakes or débitage rather than the shaping of a cobble into a predetermined morphology ([@inizan1999], [@duke2021]).

The origins of shaping strategies are most commonly linked to the Acheulean, particularly those exemplified by bifacial shaping strategies on handaxes. These strategies gradually became more refined, with mature bifacial shaping marked by the concept of bifacial symmetry, thinning and edge refinement, and the achievement of a standardized final shape ([@beyene2013]). Compared to flaking technology in Oldowan, these processes demand greater cognitive planning and precise manual control ([@wynn1995], [@stout2015]). Although the spheroids from QSY do not qualify as Large Cutting Tools (LCTs), nor does the site yield other typically Acheulean products, the faceted spheroids nonetheless exhibit a level of technological and morphological complexity comparable to that of LCTs. They also share key features, including a hierarchical and standardized reduction sequence and a high degree of symmetry and surface regularity in the final product.

These findings also contribute to the long-standing debate over the nature of spheroids—whether they should be interpreted as byproducts of flake production (exhausted cores), or as targeted end products of shaping strategies. In the case of QSY facted spheroids, the evidence strongly supports the latter. Faceted spheroids recovered from other Early Pleistocene sites in anBarranco Leo ́nd and ‘Ubeidiya have also been reported to show evidence consistent with shaping strategy ([@Muller2023], [@tittonSubspheroidsLithicAssemblage2020]). These intentionally shaped spheroids appear contemporaneous with, or even predate, the emergence of Acheulean technology. It is important to note, however, that many spheroids from East Africa may not fall into this category. As noted by [@mora2005], these quartz-based spheroids may have been produced through percussive techniques. They are often covered with battering marks, exhibit smoother surfaces, and lack ridges—features that suggest they may be better classified as bolas within a typological framework ([@leakey1971]).

Another issue we aim to address is the relationship between polyhedrons, spheroids, and bolas. While no bolas were identified at the QSY site, our quantitative analyses clearly differentiate polyhedrons from spheroids. These two artifact types differ in size, morphology, and production sequence. Polyhedrons at QSY represent heavily reduced cores, whose edge angle increase results from extensive flake removal. Their primary purpose lies in producing flake, placing them within a flaking strategy. In contrast, spheroids were intentionally shaped as end products, reflecting a shaping strategy. These two technological approaches of flaking and shaping strategies coexisted at QSY, indicating a diversity of reduction goals.

However, we found that although the morphology and reduction technology of spheroids and polyhedrons can be clearly distinguished, spheroids in early phases may be difficult to differentiate from cores. In their initial phases of reduction, spheroids often resemble cores due to similar flake removal patterns. At this stage, they may be classified as cores, especially since the flakes produced during the initial shaping are themselves usable. However, it remains unclear whether this typological transformation was intentional or unplanned. In addition, the spheroids at QSY appear to have undergone another form of typological transformation—from hammerstones to spheroids. one specimen initially served as percussive tools but were later subjected to flake removal, suggesting a shift in function from battering to shaping.

## The Technological and Typological Variability of Spheroids in China
